One-arm spirals in self-gravitating dead zones in protoplanetary discs

### 2014-10-22 paper/model.tex
minor update

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We further assume a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by $c_s=H\Omega_k$, where $H=hR$ is the
disc scale-height with constant aspect-ratio $h$ chosen below, and
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency.

>>>>

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We further assume a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by $c_s=H\Omega_k$, $H=hR$ is the
disc scale-height with constant aspect-ratio $h$ chosen below, and
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency.

### 2014-10-22 paper/paper.tex
minor update

\maketitle
\begin{abstract}
  Theoretical models of circumstellar or protoplanetary discs indicate
  the possible existence of `dead zones' --- regions in the disc
  where mass accretion is greatly reduced because of inefficient
  magneto-hydrodynamic turbulence that would otherwise enable 
  accretion. It has been suggested that mass built-up in a dead zone
  may lead to gravitational instability. However, the type of
  gravitational instability relevant in this context has not been
  clarified. In this work, we use numerical simulations to explore
  gravitational instability in astrophysical discs in which only a 
  sub-portion of it is self-gravitating, taken as a toy model for
  protoplanetary disc containing a dead zone. We conduct numerical hydrodynamic
  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
  and PLUTO codes. In all cases we observe the development of a spiral
  mode with azimuthal wavenumber $m=1$ inside the dead
  zone. Surprisingly, the co-rotation radius of the spiral is well
  outside the dead zone, resulting in a nearly stationary one-arm spiral
  pattern in the dead zone. Our simulations demonstrate the potential
  importance global gravitational interaction between a
  self-gravitating dead zone and the exterior, non-self-gravitating
  parts of the disc.  
\end{abstract}

>>>>

\maketitle
\begin{abstract}
  Theoretical models of circumstellar or protoplanetary discs indicate
  the possible existence of `dead zones' --- regions in the disc
  where mass accretion is greatly reduced because of inefficient
  magneto-hydrodynamic turbulence that would otherwise enable 
  accretion. It has been suggested that mass built-up in a dead zone
  may lead to gravitational instability. However, what type of
  gravitational instability is relevant in this context has not been
  clarified. In this work, we use numerical simulations to explore
  gravitational instability in astrophysical discs in which only a 
  sub-portion of it is self-gravitating, taken as a toy model of a 
  protoplanetary disc containing a massive dead zone. We conduct
  numerical hydrodynamic 
  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
  and PLUTO codes. In all cases we observe the development of a spiral
  mode with azimuthal wavenumber $m=1$ inside the dead
  zone. Surprisingly, the co-rotation radius of the spiral is well
  outside the dead zone, resulting in a nearly stationary one-arm spiral
  pattern in the dead zone. Our simulations demonstrate the 
  importance of global gravitational interaction between a 
  self-gravitating dead zone and the exterior, non-self-gravitating
  parts of the disc.  
\end{abstract}

### 2014-10-22 paper/results.tex
minor update

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
depend on $\phi$, and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. 
\subsection{FARGO 2D}\label{fargo_fiducial}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$ and resolution $N_R\times N_\phi = 512\times 1024$ or
about $8$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}, and 
set $M=10$ for random perturbations (Eq. \ref{randpert}).

Snapshots from the disc evolution is shown in Fig. \ref{fargo_2d} in
terms of the non-axisymmetric surface density.

@@

\subsection{ZEUS-MP 3D}

>>>>

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. 
\subsection{FARGO 2D}\label{fargo_fiducial}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$. This gives a total disc mass $M_{d}=0.086M_*$, and the
mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
We use a resolution of $N_R\times N_\phi = 512\times 1024$ or
about $8$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.
Perturbations are set initially with $M=10$ and
$\delta\in[-10^{-3},10^{-3}]$ chosen randomly but does not depend
on $\phi$.

%set $M=10$ for random perturbations (Eq. \ref{randpert}).

Snapshots from the disc evolution are shown in Fig. \ref{fargo_2d} in
terms of the non-axisymmetric surface density. At early times
$t\lesssim100P_0$ the disc is dominated by low-amplitude high-$m$
perturbations centred in the dead zone. Eventually a $m=1$ spiral 
grows and dominates inside the dead zone.

Fig. \ref{fargo_2d_angmom} compares the disc angular momenta
evolution, most of which is contained in the $m=0$ and $m=1$ modes. 
The $m=1$ spiral has an associated negative angular
momentum. Its growth is correlated with an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 \sim - \Delta
J_1$. Note that FARGO does not explicitly conserve angular
momentum. In addition, there may be an angular momentum flux due to
self-gravity across disc boundaries since our domain size is finite. 
Nevertheless, we find the total angular momentum is conserved to 
$\Delta J = O(10^{-7}J)$, and is much smaller in magnitude than the change in the
angular momenta components, $|\Delta J_m|$. 
Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
angular momentum from the $m=1$ spiral to the background disc.

@@

\subsection{ZEUS-MP 3D}
The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

Our 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$.
